Capstone Movielens Report

### 2025-05-08 reports/capstone-movielens-report.draft4.Rmd
Draft report updated.

### Taking into account User Effect
\
To improve our model let's now take into consideration user effects as explained in [Section _23.4 User effects_](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#user-effects) of the *Course Textbook*. 
If we visualize the average rating for each user the way the [the author](https://x.com/rafalab) shows, we can see that there is substantial variability in the average ratings across users: 
```{r}
hist(edx.user_mean_ratings$mean_rating, nclass = 30)
```

@@

::: {.noteblock data-latex=""}
The full source code of the _User Effect Model RMSE_ computation is available in the [Computing the RMSE taking into account User Effect](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L655) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

### Taking into account Movie effects
\

In [Section _23.5 Movie effects_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movie-effects) the author draws our attention to the fact that some movies are generally rated higher than others. He also explains that a linear model with a _treatment effect_ $\beta_j$ for each movie can be used in this case, which can be interpreted as movie effect or the difference between the average ranking for movie $j$ and the overall average $\mu$:

@@

We can now construct predictors and see how much the `RMSE` improves[@IDS2_23-5]:
```{r eval=FALSE}
user_and_movie_effects_rmse <- test_set |> 
    left_join(data.frame(userId = as.integer(names(a)), a = a), by = "userId") |>
    left_join(data.frame(movieId = as.integer(names(b)), b = b), by = "movieId") |>
    mutate(resid = rating - clamp(mu + a + b)) |>  
    filter(!is.na(resid)) |>
    pull(resid) |> rmse()

print(user_and_movie_effects_rmse)
```
### Utilizing Penalized least squares
\
[Section _23.6 Penalized least squares_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#penalized-least-squares) explains why and how we should use _Penalized least squares_ to improve our predictions. The author also explains that the general idea of penalized regression is to control the total variability of the movie effects: $\sum_{j=1}^n \beta_j^2$. Specifically, instead of minimizing the least squares equation, we minimize an equation that adds a penalty:

@@

str(b_reg)
```
Finally, let's verify that the penalized estimates $\hat{b}_i(\lambda)$ we have just computed actually result in the minimal _RMSE_ figured out above: 
```{r eval=FALSE}
reg_rmse(b_reg)
```
### Accounting for Date effects

>>>>

### User Effect Model
\
To improve our model let's now take into consideration user effects as explained in [Section _23.4 User effects_](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#user-effects) of the *Course Textbook*. 
If we visualize the average rating for each user the way the [the author](https://x.com/rafalab) shows, we can see that there is substantial variability in the average ratings across users: 
```{r}
hist(edx.user_mean_ratings$mean_rating, nclass = 30)
```

@@

::: {.noteblock data-latex=""}
The full source code of the _User Effect Model RMSE_ computation is available in the [Compute RMSE for User Effect Model](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L655) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

### User+Movie Effect (UME) Model
\

In [23.5 Movie effects](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movie-effects) section of the *Course Textbook* the author draws our attention to the fact that some movies are generally rated higher than others. He also explains that a linear model with a _treatment effect_ $\beta_j$ for each movie can be used in this case, which can be interpreted as movie effect or the difference between the average ranking for movie $j$ and the overall average $\mu$:

@@

Inspired by this idea, a few support functions were developed by the author of this report, which we will use for our further analysis.

#### UME Model: Support Functions
\

::: {.noteblock data-latex=""}
The full source code of the functions described in this section is available in the [User+Movie Effect Model Functions](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L1) section of the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script on _GitHub_.
:::

##### `train_user_movie_effect` Function
\

We use this function to build and train our model using the `train_set` dataset:
```{r eval=FALSE}
train_user_movie_effect <- function(train_set, lambda = 0){
  if (is.na(lambda)) {
    stop("Function: train_user_movie_effect
`lambda` is `NA`")
  }

  UM.effect <- train_set |>
    left_join(edx.user_effect, by = "userId") |>
    mutate(resid = rating - (mu + a)) |> 
    group_by(movieId) |>
    summarise(b = mean_reg(resid, lambda), n = n())

  stopifnot(!is.na(mean(UM.effect$b)))
  UM.effect
}
```

::: {.noteblock data-latex=""}
The function described above accepts the `lambda` parameter, which we will need later for the _Regularization_ method. We also use the [mean_reg](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L63) function call, which we will also need for the *Regularization*.
We will explain that later in the [Regularization Method] section. 
For now, we omit the `lambda` parameter, accepting its default value `lambda = 0`. In this case, the `mean_reg` function is equivalent to the simple `mean` one.
:::

```{r eval=FALSE}
## Regularization --------------------------------------------------------------
mean_reg <- function(vals, lambda = 0, na.rm = TRUE){
  if (is.na(lambda)) {
    stop("Function: mean_reg
`lambda` is `NA`")
  }

  names(lambda) <- NULL
  sums <- sum(vals, na.rm = na.rm)
  N <- ifelse(na.rm, sum(!is.na(vals)), length(vals))
  sums/(N + lambda)
}
```

##### `train_user_movie_effect.cv` Function
\

We use the [train_user_movie_effect.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L17) function to build and train our model using the `5-Fold Cross Validation` method. Below, we provide the most important part of the code of that function:
```{r eval=FALSE}
train_user_movie_effect.cv <- function(lambda = 0){
# ...
  start <- put_start_date()
  user_movie_effects_ls <- lapply(edx_CV, function(cv_fold_dat){
    cv_fold_dat$train_set |> train_user_movie_effect(lambda)
  })
  put_end_date(start)
  put_log("Function: train_user_movie_effect.cv:
User+Movie Effect list have been computed")

  user_movie_effects_united <- union_cv_results(user_movie_effects_ls)

  user_movie_effect <- user_movie_effects_united |>
    group_by(movieId) |>
    summarise(b = mean(b), n = mean(n))
  # ...
  user_movie_effect
}

::: {.noteblock data-latex=""}
Here we use the function call [union_cv_results](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L432), which is defined in the script [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R), to aggregate the "cross-validation" results.
:::
```{r eval=FALSE}
union_cv_results <- function(data_list) {
  out_dat <- data_list[[1]]

  for (i in 2:CVFolds_N){
    out_dat <- union(out_dat, 
                     data_list[[i]])
  }

  out_dat
}
```

##### `calc_user_movie_effect_MSE` Function
\

The code of the function [calc_user_movie_effect_MSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L55) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `MSE` of the `UME Model` for the given `Test Set` is provided below:
```{r eval=FALSE}
calc_user_movie_effect_MSE <- function(test_set, um_effect){
  mse.result <- test_set |>
    left_join(edx.user_effect, by = "userId") |>
    left_join(um_effect, by = "movieId") |>
    mutate(resid = rating - clamp(mu + a + b)) |> 
    pull(resid) |> mse()

  stopifnot(!is.na(mse.result))
  mse.result
}
```

##### `calc_user_movie_effect_MSE.cv` Function
\

The code of the function [calc_user_movie_effect_MSE.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L72) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `5-Fold Cross Validation` MSE result of the `UME Model` is provided below:
```{r eval=FALSE}
calc_user_movie_effect_MSE.cv <- function(um_effect){
  put_log("Function: user_movie_effects_MSE.cv:
Computing the RMSE taking into account User+Movie Effects...")
  start <- put_start_date()
  user_movie_effects_MSEs <- sapply(edx_CV, function(cv_fold_dat){
    cv_fold_dat$validation_set |> calc_user_movie_effect_MSE(um_effect)
  })
  put_end_date(start)

  put_log1("Function: user_movie_effects_MSE.cv:
MSE values have been plotted for the %1-Fold Cross Validation samples.", 
           CVFolds_N)

  mean(user_movie_effects_MSEs)
}

##### `calc_user_movie_effect_RMSE` Function
\

The code of the function [calc_user_movie_effect_RMSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L51) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `RMSE` of the `UME Model` for the given `Test Set` is provided below:

```{r eval=FALSE}
calc_user_movie_effect_RMSE <- function(test_set, um_effect){
  mse <- test_set |> calc_user_movie_effect_MSE(um_effect)
  sqrt(mse)
}
```

##### `calc_user_movie_effect_RMSE.cv` Function
\

The code of the function [calc_user_movie_effect_RMSE.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L65) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `5-Fold Cross Validation` RMSE result of the `UME Model` is provided below:

```{r eval=FALSE}
calc_user_movie_effect_RMSE.cv <- function(um_effect){
  user_movie_effects_MSE <- calc_user_movie_effect_MSE.cv(um_effect)
  um_effect_RMSE <- sqrt(user_movie_effects_MSE)
  put_log2("Function: user_movie_effects_RMSE.cv:
%1-Fold Cross Validation ultimate RMSE: %2", CVFolds_N, um_effect_RMSE)
  um_effect_RMSE
}

#### Model Building
\

::: {.noteblock data-latex=""}
The full source code of builing and training the current model is available in the [Model building: User+Movie Effect](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L721) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

Below, we provide the most important code of the training our model using the `5-Fold Cross Validation` method:
```{r eval=FALSE}
  cv.UM_effect <- train_user_movie_effect.cv()
```
```{r}
par(cex = 0.7)
hist(cv.UM_effect$b, 30, xlab = TeX(r'[$\hat{beta}_{j}$)]'),
     main = TeX(r'[Histogram of $\hat{beta}_{j}$]'))
```

We can now construct predictors and see how much the `RMSE` improves[@IDS2_23-5]:
```{r eval=FALSE}
cv.UM_effect.RMSE <- calc_user_movie_effect_RMSE.cv(cv.UM_effect)

RMSEs.ResultTibble.UME <- RMSEs.ResultTibble.UE |> 
  RMSEs.AddRow("User+Movie Effect Model", cv.UM_effect.RMSE)

```
```{r}
RMSE_kable(RMSEs.ResultTibble.UME)
```

### Utilizing Penalized least squares
\
[Section _23.6 Penalized least squares_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#penalized-least-squares) explains why and how we should use _Penalized least squares_ to improve our predictions. The author also explains that the general idea of penalized regression is to control the total variability of the movie effects: $\sum_{j=1}^n \beta_j^2$. Specifically, instead of minimizing the least squares equation, we minimize an equation that adds a penalty:

@@

str(b_reg)
```
Finally, let's verify that the penalized estimates $\hat{b}_i(\lambda)$ we have just computed actually result in the minimal _RMSE_ figured out above: 
```{r eval=FALSE}
reg_rmse(b_reg)
```

##### `` Function
\

##### `` Function
\

##### `` Function
\

### Accounting for Date effects
